---
title: "STA2023 Review:<br>Describing Data"
execute:
  echo: true
  warning: false
  message: false
  error: true
format: 
  revealjs:
    theme: default
    css: style_lecture.css
    self-contained: true
    slide-number: true
    footer: "STA4173 - Biostatistics - Fall 2026"
    width: 1600
    height: 900
    df-print: paged
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
editor: source
---

## Introduction: Topics

- Basic descriptives:
    - Continuous variables:
        - Mean
        - Median
        - Variance and standard deviation
        - Range and interquartile range
    - Categorical variables:
        - Count
        - Overall percentage
        - Row percentage
        - Column percentage
        
## Introduction: Data

- We will be using data from the kingdom of [Equestria](https://en.wikipedia.org/wiki/Equestria) (yes, from [My Little Pony](https://en.wikipedia.org/wiki/My_Little_Pony:_Friendship_Is_Magic)).

- [Mane Six](https://equestripedia.org/wiki/Mane_Six):
    - Twilight Sparkle (Unicorn $\to$ Alicorn)
    - Applejack (Earth Pony)
    - Fluttershy (Pegasus)
    - Pinkie Pie (Earth Pony)
    - Rainbow Dash (Pegasus)
    - Rarity (Unicorn)

## Introduction: Data

```{r}
#| echo: false
library(tidyverse)
library(gsheet)
library(DT)
library(janitor)
library(ssstats)
mlp_data <- gsheet2tbl("https://docs.google.com/spreadsheets/d/1fgkPVXXmif9sXP1Gg7BEeRfKO3tayn1vvb2ep7eQqjs/edit?gid=0#gid=0") 
mlp_data %>% head(5)
```

## Types of Variables: Qualitative

::: {.notebook-entry}
::: {.notebook-tab}
Categorical variable
:::
A **qualitative** or **categorical** variable classifies an observation into one of two or more groups or categories.


- Nominal: purely qualitative and unordered

- Ordinal: data can be ranked, but intervals between ranks may not be equivalent

:::
    
- Examples:
    - favorite color
    - education level
    - blood type
    
## Types of Variables: Quantitative    

::: {.notebook-entry}
::: {.notebook-tab}
Continuous variable
:::
A **quantitative** or **continuous** variable takes numerical values for which arithmetic operations such as adding and averaging make sense; typically has a unit of measure. 

- Interval: meaningful differences between values, but no true zero point

- Ratio: meaningful differences and a true zero point

:::

- Examples:
    - age (years)
    - temperature (Celsius)
    - daily hours of sleep

## Types of Variables: Example

```{r}
#| echo: false
mlp_data <- gsheet2tbl("https://docs.google.com/spreadsheets/d/1fgkPVXXmif9sXP1Gg7BEeRfKO3tayn1vvb2ep7eQqjs/edit?gid=0#gid=0")
datatable(mlp_data,
          options = list(pageLength = 3, # how many obs shown on slide
                 scrollX = FALSE, # can scroll across cols
                 lengthChange = FALSE, # cannot change number of rows shown
                 info = FALSE, # no info about number of obs
                 searching = FALSE, # no search bar
                 ordering = FALSE, # no ordering
                 dom = 't'),  # just display the table
          rownames = FALSE) # do not include row number
```

- Name: the pony's name
- Type: type of pony (Earth, Pegasus, Unicorn, Alicorn)
- Sex: sex/age of pony (Coal, Filly, Stallion, Mare)
- Flying speed: average flying speed (km/hr) for winged ponies
- Friendship: a harmony index from friendship activities (0-10)
- Magical energy: measured magical energy output (sparkles) for magical ponies
- Tail shimmer: how much light reflected by the pony's tail (lux)

## Describing Data: Why?

- Why do we describe data? We want to tell a story!
    - Summarize *n* observations into a single description
    - Understand what is in the data
    - Spot patterns, missing data, or outliers
    - Compare groups or spot differences or oddities
    
## Describing Data: How?    

- How do we describe data?
    - Numbers
        - Frequency table
        - Mean & standard deviation
        - Median & IQR
    - Graphs
        - Bar charts
        - Box plots
        - Histograms
  
## Point Estimation: Mean

::: {.notebook-entry}
::: {.notebook-tab}
Mean
:::
The **mean** is the average of a set of values and describes the *center* of the data distribution.

$$
\bar{y} = \frac{\sum_{i=1}^n y_i}{n}
$$

:::

## Point Estimation: Mean

- Find the mean for the flying speeds (km/hr) of 5 ponies: {10, 20, 30, 40, 100}

$$
\bar{y} = \frac{\sum_{i=1}^n y_i}{n} = \frac{10 + 20 + 30 + 40 + 100}{5} = 40
$$

- The average flying speed for winged ponies is 40 km/hr.

## Point Estimation: Median

::: {.notebook-entry}
::: {.notebook-tab}
Median
:::
The **median** is the middle value in an ordered dataset and describes the *center* of the data distribution.

- Odd number of observations: the median is the middle value.
- Even number of observations: the median is the average of the middle two values.

:::

## Point Estimation: Median

- Find the median for the flying speeds (km/hr) of 5 ponies: {10, 20, 30, 40, 100}
    - First, we sort the data: {10, 20, 30, 40, 100}
    - Then, find the middle number: 30
    
- The median flying speed for winged ponies is 30 km/hr.    

## Point Estimation: Variance

::: {.notebook-entry}
::: {.notebook-tab}
Variance
:::
The **variance** is the average of squared differences from the mean and describes the *spread* of the data distribution.

$$
s_y^2 = \frac{\sum_iy_i^2 - (\sum_iy_i)^2/n}{n-1}
$$

:::

## Point Estimation: Standard Deviation

::: {.notebook-entry}
::: {.notebook-tab}
Standard deviation
:::
The **standard deviation** is the average distance from the mean and describes the *spread* of the data distribution.

$$
s_y = \sqrt{s_y^2}
$$

:::

## Point Estimation: Variance and Standard Deviation
    
- Find the variance and standard deviation for the flying speeds (km/hr) of 5 ponies: {10, 20, 30, 40, 100}    

$$
\begin{align*}
s_y^2 &= \frac{\sum_iy_i^2 - (\sum_iy_i)^2/n}{n-1}  \\
&= \frac{(10^2+...+100^2)-(10+...+100)^2/5}{4} \\
&= 1250
\end{align*}
$$

- The variance is 1250 (km/hr)^2^

## Point Estimation: Variance and Standard Deviation
    
- Find the variance and standard deviation for the flying speeds (km/hr) of 5 ponies: {10, 20, 30, 40, 100}    

- The variance is 1250 (km/hr)^2^

$$
s_y = \sqrt{s^2_y} = \sqrt{1250} \approx 35.36
$$

- The standard deviation is 35.36 km/hr.

## Point Estimation: Range

::: {.notebook-entry}
::: {.notebook-tab}
Range
:::
The **range** is the difference between the maximum and minimum values.

$$
\text{range} = \text{max}(y) - \text{min}(y)
$$

:::

## Point Estimation: Range

- Find the range for the flying speeds (km/hr) of 5 ponies: {10, 20, 30, 40, 100} 

$$
\begin{align*}
\text{range} = \text{max}(y) - \text{min}(y) = 100 - 10 = 90
\end{align*}
$$

- The range of the flying speeds is 90 km/hr.

## Point Estimation: Interquartile Range

::: {.notebook-entry}
::: {.notebook-tab}
Interquartile Range
:::
The **interquartile range** (IQR) is the difference between 25^th^ percentile and the 75^th^ percentile and describes the *spread* of the middle 50% of the data.

$$
\text{IQR} = \text{P}_{75} − \text{P}_{25}
$$

:::

## Point Estimation: Interquartile Range

- Find the IQR for the flying speeds (km/hr) of 5 ponies: {10, 20, 30, 40, 100} 
    - Recall that the median is 30.
    - We then find $P_{25}$ using {10, 20} and $P_{75}$ using {40, 100} 
    - Thus, $P_{25} = 15$ and $P_{75} = 70$.
    
$$
\begin{align*}
\text{IQR} = \text{P}_{75} − \text{P}_{25} = 70 - 15 = 55
\end{align*}
$$ 

- The IQR of the flying speeds is 55 km/hr.

## Point Estimation: Proportion

::: {.notebook-entry}
::: {.notebook-tab}
Proportion
:::
The **proportion** is a type of mean for categorical data

$$
\hat{p} = \frac{\sum_{i=1}^n y_i}{n},
$$

:::

- Note that in this case, 

$$
y_i = 
\begin{cases}
  1 & \text{if in category }i \\
  0 & \text{otherwise}
\end{cases}
$$

## Point Estimation: Proportion

- Find the proportion of ponies that have wings in the following sample: {Y, N, Y, Y, N, Y} 

- Count the number of "Y" responses and divide by total:

$$
\hat{p} = \frac{\sum_{i=1}^n y_i}{n} = \frac{4}{6} \approx 0.67
$$

- The proportion of ponies with wings is 0.667 (or 66.7%).

## Point Estimation: Frequency Table

::: {.notebook-entry}
::: {.notebook-tab}
Frequency table
:::
A **frequency table** is a table showing how often each value appears in a dataset.

- For each category, $i$, we report $n_i$ ($\%_i$)

:::

## Point Estimation: Frequency Table

- Find the frequency table for the following sample of 8 ponies:  {Earth, Pegasus, Unicorn, Earth, Pegasus, Pegasus, Unicorn, Alicorn}

:::{.columns}
::: {.column width="40%"}
- Frequencies:
    - Alicorn: $n_{\text{A}} = 1$
    - Earth: $n_{\text{E}} = 2$
    - Pegasus: $n_{\text{P}} = 3$
    - Unicorn: $n_{\text{U}} = 2$
    
:::

::: {.column width="50%"}
- Proportions:
    - Alicorn: $\hat{p}_{\text{A}} = 1/8 = 0.125$
    - Earth: $\hat{p}_{\text{E}} = 2/8 = 0.250$
    - Pegasus: $\hat{p}_{\text{P}} = 3/8 = 0.375$
    - Unicorn: $\hat{p}_{\text{U}} = 2/8 = 0.250$ 
    
:::
:::

## Point Estimation: Frequency Table

- Putting this into a table,

```{r}
#| echo: false
labels <- c("Pony Type", "n (%)")
tibble(pony_type = haven::labelled(c("Alicorn", "Earth", "Pegasus", "Unicorn"),
                                   label = labels[1]),
                        n_pct = haven::labelled(c("1 (12.5%)", "2 (25.0%)", "3 (37.5%)", "2 (25.0%)"), 
                                                label = labels[2])) %>% 
  datatable(options = list(pageLength = 4,
                           scrollX = FALSE,
                           lengthChange = FALSE,
                           info = FALSE,
                           searching = FALSE,
                           ordering = FALSE,
                           dom = 't'),
            rownames = FALSE)
```

## Point Estimation: Contingency Table

::: {.notebook-entry}
::: {.notebook-tab}
Contingency table
:::
A **contingency table** is a frequency table that simultaneously summarizes two qualitative variables.
:::

- This is often referred to as an $R \times C$ contingency table.

## Point Estimation: Contingency Table

- We will not concern ourselves with the derivation, but will rely on R.

- Consider this data,

```{r}
#| echo: false
tibble(pony_id = LETTERS[1:8],
       pony_type = c("Earth", "Pegasus", "Unicorn", "Pegasus", "Alicorn", "Earth", "Pegasus", "Unicorn"),
       flying_status = c("No", "Yes", "No", "Yes", "Yes", "No", "Yes", "No")
) %>%
    datatable(options = list(pageLength = 4,
                             scrollX = FALSE,
                             lengthChange = FALSE,
                             info = FALSE,
                             searching = FALSE,
                             ordering = FALSE),
              rownames = FALSE)
```

## Point Estimation: Contingency Table

- The resulting contingency table would look something like this:
    - We are using *column totals* as our denominators.
    
```{r}
#| echo: false
tibble(pony_id = LETTERS[1:8],
       pony_type = c("Earth", "Pegasus", "Unicorn", "Pegasus", "Alicorn", "Earth", "Pegasus", "Unicorn"),
       flying_status = c("No", "Yes", "No", "Yes", "Yes", "No", "Yes", "No")
) %>%
  n_pct(pony_type, flying_status)
```

## Graphs: Box Plots

- Box plots display the distribution of a continuous variable using the five number summary; this gives us an idea about the *spread* and *skewness* of the data:
    - Whisker: Minimum
    - Beginning of box: 25^th^ percentile (first quartile; Q1, P25)
    - "Middle" of box: Median (50th percentile, second quartile; Q2, P50)
    - End of box: 75^th^ percentile (third quartile; Q3, P75)
    - Whisker: Maximum
    
- Note: there are different ways to define the whiskers.
    - I use the min/max as whiskers when sketching by hand.
    - `ggplot()` uses 1.75 $\times$ IQR.
    
## Graphs: Box Plots

- Describe this box plot:

```{r}
#| fig-align: 'center'
#| fig-alt: 'Box plot of the number of apples eaten by a sample of ponies.'
#| echo: false
tibble(x = c(10, 11, 11, 12, 12, 12, 13, 13, 14)) %>% 
  ggplot(aes(x = x)) +
  geom_boxplot(fill = "#F9E79F") + 
  labs(x = "Number of Apples Eaten") +
  theme_minimal(base_size = 20) +
  theme(axis.ticks.y = element_blank()) 
```

## Graphs: Box Plots

- Describe this box plot:

```{r}
#| echo: false
#| fig-align: 'center'
#| fig-alt: 'Box plot of the number of apples eaten by a sample of ponies.'
tibble(x = c(5, 6, 6, 7, 7, 8, 9, 15, 20)) %>% 
  ggplot(aes(x = x)) +
  geom_boxplot(fill = "#A00000") + 
  labs(x = "Number of Apples Eaten") +
  theme_minimal(base_size = 20) + 
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) 
```

## Graphs: Box Plots

- Describe this box plot:

```{r}
#| echo: false
#| fig-align: 'center'
#| fig-alt: 'Box plot of the number of apples eaten by a sample of ponies.'
tibble(x = c(1, 2, 3, 4, 5, 5, 5, 6, 6, 6)) %>% 
  ggplot(aes(x = x)) +
  geom_boxplot(fill = "#FFD7B5") + 
  labs(x = "Number of Apples Eaten") +
  theme_minimal(base_size = 20) + 
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) 
```

## Graphs: Box Plots

- Describe this box plot:

```{r}
#| echo: false
#| fig-align: 'center'
#| fig-alt: 'Box plot of the number of apples eaten by a sample of ponies.'
tibble(count = c(20, 21, 22, 22, 23, 24, 25, 25, 26,
                 10, 15, 18, 22, 23, 24, 28, 32, 35),
       spread = c(rep("Adults",9), 
                  rep("Juveniles",9))) %>% 
  ggplot(aes(x = spread, y = count, fill = spread)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Adults" = "#A8D03B", "Juveniles" = "#F26C4F")) +
  labs(y = "Number of Apples Eaten",
       x = "Pony Age Group") +
  theme_minimal(base_size = 20) +
  theme(legend.position = "none")
```


## Graphs: Histograms

- Histograms show the distribution (or *shape*) of a continuous variable.

- Values are grouped into intervals ("bins"), then the bin height demonstrates how many values fall into that interval.

- This allows us to quickly see if there are any oddities.

    - Increased proportion of a specific value/bin.

    - Any values that are "out in the tail"?

## Graphs: Histograms

```{r}
#| echo: false
set.seed(18662)
df <- bind_rows(tibble(value = rnorm(100, mean = 50, sd = 10), type = "Symmetric"),
                tibble(value = rexp(100, rate = 1/10) + 30, type = "Right Skewed"),
                tibble(value = 80 - rexp(100, rate = 1/10), type = "Left Skewed"),
                tibble(value = c(rnorm(50, 40, 5), rnorm(50, 70, 5)), type = "Bimodal"),
                tibble(value = runif(100, min = 30, max = 70), type = "Uniform"))
```

- Describe the histogram:

```{r}
#| echo: false
#| fig-align: 'center'
#| fig-alt: 'Histogram of the magic ability scores for a sample of ponies.'
df %>% 
  filter(type == "Symmetric") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 15, fill = "#FF2C8B", color = "#4A145D") +
  theme_minimal(base_size = 20) +
  labs(y = "Number of Ponies",
       x = "Magic Ability Score")
```

## Graphs: Histograms

- Describe the histogram:

```{r}
#| echo: false
#| fig-align: 'center'
#| fig-alt: 'Histogram of the magic ability scores for a sample of ponies.'
df %>% 
  filter(type == "Left Skewed") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 15, fill = "#88CCF1", color = "#0072CE") +
  theme_minimal(base_size = 20) +
  labs(y = "Number of Ponies",
       x = "Magic Ability Score")
```

## Graphs: Histograms

- Describe the histogram:

```{r}
#| echo: false
#| fig-align: 'center'
#| fig-alt: 'Histogram of the magic ability scores for a sample of ponies.'
df %>% 
  filter(type == "Uniform") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 15, fill = "#F4A460", color = "#5C3A21") +
  theme_minimal(base_size = 20) +
  labs(y = "Number of Ponies",
       x = "Magic Ability Score")
```

## Graphs: Histograms

- Describe the histogram:

```{r}
#| echo: false
#| fig-align: 'center'
#| fig-alt: 'Histogram of the magic ability scores for a sample of ponies.'
df %>% 
  filter(type == "Right Skewed") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 15, fill = "#8B5FBF", color = "#5C2D91") +
  theme_minimal(base_size = 20) +
  labs(y = "Number of Ponies",
       x = "Magic Ability Score")
```

## Graphs: Histograms

- Describe the histogram:

```{r}
#| echo: false
#| fig-align: 'center'
#| fig-alt: 'Histogram of the magic ability scores for a sample of ponies.'
df %>% 
  filter(type == "Bimodal") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 15, fill = "#FFF1A8", color = "#F7BFD8") +
  theme_minimal(base_size = 20) +
  labs(y = "Number of Ponies",
       x = "Magic Ability Score")
```

## Graphs: Bar Graphs

- Bar graphs display the distribution of categorical data.
    - The frequency or proportion of observations is displayed on the bar graph.
    
- Bar graphs usually have categories on the *x*-axis and counts or proportions on the *y*-axis.
    - Note that we could flip the axes to create a vertical bar graph.
    
- Note that the bars are *separated* on the *x*-axis to indicate the lack of continuity.

## Graphs: Bar Graphs

- Consider the bar graph, below.

```{r}
#| echo: false
#| fig-align: 'center'
#| fig-alt: 'Histogram of the magic ability scores for a sample of ponies.'
df <- tibble(fruit = c("Apple", "Banana", "Cherry", "Plum"),
             count = c(23, 17, 35, 10))

df %>% ggplot(aes(x = fruit, y = count)) +
  geom_col(fill = "#F9A602", color = "#5C3A21") +
  labs(x = "Fruit",
       y = "Number of Ponies") +
  theme_minimal(base_size = 20) 
```

## Graphs: Side-by-Side Bar Graphs

- Consider the bar graph, below.


```{r}
#| echo: false
#| fig-align: 'center'
#| fig-alt: 'Histogram of the magic ability scores for a sample of ponies.'
tibble(FavoriteFruit = c("Apples", "Bananas", "Cherries", "Grapes", "Mangoes", "Pears"),
       Earth = c(6, 6, 2, 6, 2, 6),
       Pegasus = c(6, 3, 4, 2, 3, 4)) %>%
  pivot_longer(cols = c(Earth, Pegasus), 
               names_to = "Type", 
               values_to = "Count") %>%
  ggplot(aes(x = FavoriteFruit, 
             y = Count, 
             fill = Type)) +
  geom_col(position = "dodge", 
           color = "black") +
  labs(x = "Favorite Fruit",
       y = "Number of Ponies",
       fill = "Pony Type") +
  scale_fill_manual(values = c("Earth" = "#A0522D",
                               "Pegasus" = "#87CEEB")) + 
  theme_minimal(base_size = 20)
```


## Graphs: Stacked Bar Graphs

- Consider the bar graph, below.


```{r}
#| echo: false
#| fig-align: 'center'
#| fig-alt: 'Histogram of the magic ability scores for a sample of ponies.'
set.seed(10576)
groups <- c("Earth", "Pegasus", "Unicorn")
fruits <- c("Apples", "Bananas", "Cherries", "Grapes", "Pears", "Mangoes")
pony_data <- tibble(Pony = paste0("Pony_", 1:300),
                    Group = rep(groups, each = 100),
                    FavoriteFruit = sample(fruits,
                                           size = 300,
                                           replace = TRUE,
                                           prob = c(0.25, 0.20, 0.15, 0.15, 0.15, 0.10))) 
pony_data %>%
  count(Group, FavoriteFruit) %>%
  ggplot(aes(x = Group, y = n, fill = FavoriteFruit)) +
  geom_col(position = "stack", color = "black") +
  labs(x = "Pony Type",
       y = "Number of Ponies",
       fill = "Favorite Fruit") +
  scale_fill_manual(values = c("Apples" = "#ED1C24",
                               "Bananas" = "#FFF200",
                               "Cherries" = "#F7941D",
                               "Grapes" = "#00A651",
                               "Pears" = "#662D91",
                               "Mangoes" = "#0072CE")) +
  theme_minimal(base_size = 20) 
```

## Graphs: Histograms vs Bar Graphs

- We have now reviewed two "bar style" graphs that we see regularly: histograms and bar graphs.

- We use *histograms* to see the distribution of continuous variables.
    - The *x*-axis represents numeric intervals.
    - The bars touch each other to represent continuity.

- We use *bar graphs* to see the distribution of categorical variables.
    - The *x*-axis represents categories.
    - The bars do not touch each other, implying distinct categories.

## Graphs: Scatterplots

- Scatterplots allow us to look at the relationship between two continuous variables.
    - Each point on the graph represents one observation.

- What statisticians use scatterplots for:
    - Explore patterns (aka trends or  relationships).
        - Linear relationships.
        - Non-linear relationships.
    - Detect clusters of observations.
    - Find oddities in the data (outliers).

- When we describe the relationship, we are really answering the question, "As *x* increases, what happens to *y*?"

## Graphs: Scatterplots

- Consider the scatterplot, below.

```{r}
#| echo: false
#| fig-align: 'center'
#| fig-alt: 'Scatterplot of wing power and flight speed for a sample of ponies.'
set.seed(672166)
tibble(Pony = paste0("Pegasus_", 1:30),
       WingPower = rnorm(30, mean = 700, sd = 100),
       FlightSpeed = rnorm(30, mean = 80, sd = 10)) %>%
  mutate(FlightSpeed = FlightSpeed + 0.05 * (WingPower - mean(WingPower))) %>%
  ggplot(aes(x = WingPower, y = FlightSpeed)) +
  geom_point(size = 2) +
  labs(x = "Wing Power",
       y = "Flight Speed (km/h)") +
  theme_minimal(base_size = 20) 
```

## Graphs: Scatterplots

- Consider the scatterplot, below.

```{r}
#| echo: false
#| fig-align: 'center'
#| fig-alt: 'Scatterplot of hours worked and apples bucked for a sample of ponies.'
set.seed(561187)
tibble(Pony = paste0("Earth_", 1:30),
       HoursWorked = rnorm(30, mean = 6, sd = 1.5),
       ApplesBucked = rnorm(30, mean = 300, sd = 50)) %>%
  mutate(ApplesBucked = ApplesBucked + 25 * (HoursWorked - mean(HoursWorked)))  %>%
  ggplot(aes(x = HoursWorked, y = ApplesBucked)) +
  geom_point(size = 2) +
  labs(x = "Hours Worked",
       y = "Apples Bucked") +
  theme_minimal(base_size = 20) 
```

## Graphs: Scatterplots

- Consider the scatterplot, below.

```{r}
#| echo: false
#| fig-align: 'center'
#| fig-alt: 'Scatterplot of hours worked and apples bucked for a sample of ponies.'
set.seed(561187)
tibble(Pony = paste0("Earth_", 1:30),
       HoursWorked = rnorm(30, mean = 6, sd = 1.5),
       ApplesBucked = rnorm(30, mean = 300, sd = 50)) %>%
  mutate(ApplesBucked = ApplesBucked + 25 * (HoursWorked - mean(HoursWorked)))  %>%
  ggplot(aes(x = HoursWorked, y = ApplesBucked)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Hours Worked",
       y = "Apples Bucked") +
  theme_minimal(base_size = 20) 
```

## Graphs: Scatterplots

- Consider the scatterplot, below.

```{r}
#| echo: false
#| fig-align: 'center'
#| fig-alt: 'Scatterplot of hours of magic practice and spell accuracy for a sample of ponies.'
set.seed(9876)
tibble(HoursPracticed = seq(0, 10, length.out = 50),
       SpellAccuracy = 60 + 20 * HoursPracticed - 2 * HoursPracticed^2 + rnorm(50, sd = 5)) %>%
  ggplot(aes(x = HoursPracticed, y = SpellAccuracy)) +
  geom_point(size = 2) +
  labs(x = "Hours of Magic Practice",
       y = "Spell Accuracy (%)") +
  theme_minimal(base_size = 20) 
```

## Graphs: Scatterplots

- Consider the scatterplot, below.

```{r}
#| echo: false
#| fig-align: 'center'
#| fig-alt: 'Scatterplot of hours of magic practice and spell accuracy for a sample of ponies.'
set.seed(9876)
tibble(HoursPracticed = seq(0, 10, length.out = 50),
       SpellAccuracy = 60 + 20 * HoursPracticed - 2 * HoursPracticed^2 + rnorm(50, sd = 5)) %>%
  ggplot(aes(x = HoursPracticed, y = SpellAccuracy)) +
  geom_point(size = 2) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Hours of Magic Practice",
       y = "Spell Accuracy (%)") +
  theme_minimal(base_size = 20) 
```

## Wrap Up

- We have covered ("reminded" ourselves of) a lot today!
    - Always remember that I do not expect you to:
        - Memorize code.
        - Produce code in a timed environment.
        - Automatically know how to do these things.
    - I do expect you to:
        - Use your resources (lecture slides, GitHub website).
        - Try your best.
        
## Wrap Up        

- Today's lecture:
    - Basic summarization of data.
    - Basic data visualization.
    
- Wednesday's lecture:   
    - Review of statistical inference.
        - Confidence intervals.
        - Hypothesis tests.



    
