---
title: "Two-Sample Means"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    self-contained: true
    slide-number: true
    center: false
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    incremental: false
editor: source
---
        
## Introduction: Topics

- We previously have reviewed estimation and basic statistical inference.
    - Point estimation
    - Confidence intervals
    - Hypothesis testing
    
- Today we will focus on the following scenarios today:
    - Two-sample means, independent data ($\mu_1-\mu_2$)
    - Two-sample means, dependent data ($\mu_d$)


## Introduction: Data

- We are using data from the kingdom of [Equestria](https://en.wikipedia.org/wiki/Equestria) from [My Little Pony](https://en.wikipedia.org/wiki/My_Little_Pony:_Friendship_Is_Magic).

- [Mane Six](https://equestripedia.org/wiki/Mane_Six):
    - Twilight Sparkle (Unicorn $\to$ Alicorn)
    - Applejack (Earth Pony)
    - Fluttershy (Pegasus)
    - Pinkie Pie (Earth Pony)
    - Rainbow Dash (Pegasus)
    - Rarity (Unicorn)

```{r}
#| echo: false
library(ssstats)
library(tidyverse)

set.seed(8651)
wing_flap <- tibble(pony_names = c("Rainbow Dash", "Soarin", "Spitfire", "Cloudchaser", "Fleetfoot", "Lightning Dust", "Windy Whistles", "Stormy Skies", "Gusty Gale", "Zephyr Breeze", "Skydancer", "Aero Ace", "Nimbus Sprint", "Breeze Wing", "Flash Fly", "Twister Tail", "Sirensong", "Aero Wave", "Sky Dancer", "Gale Gust", "Storm Strider", "Cyclone Swift", "Wind Whirl", "Jet Stream", "Flurry Flap"),
                    wing_flap_rate = round(rnorm(n = length(pony_names), mean = 50, sd = 8),1),
                    target = if_else(wing_flap_rate>=50, "Above", "Below"),
                    apples = if_else(target == "Above", round(rnorm(n = length(pony_names), mean = 25, sd = 2)), round(rnorm(n = length(pony_names), mean = 15, sd = 3))),
                    pre_training_wfr = round(rnorm(n = length(pony_names), mean = 50, sd = 8),1),
                    post_training_wfr =round(rnorm(n = length(pony_names), mean = 50, sd = 4),1))
```

    
## Independent Data

- **Independent data:** Observations in one group (or sample) do not influence or relate to observations in another group.
    - Examples:
        - Comparing the cruising speeds of a random sample of Pegasi vs. a random sample of Unicorns flying a short course.
        - Measuring friendship lesson quiz scores for a group of Cutie Mark Crusaders vs. a group of Wonderbolts Cadets.
        - Examining the graduation rates between Unicorns and Pegasi.
        
## Dependent Data        

- **Dependent (paired) data:** Each observation in the first sample is paired with exactly one observation in the second sample. 
  - Examples:
      - Students' magic‐proficiency scores before and after Princess Celestia's advanced spell workshop.
      - Applejack's apple‐yield (in bushels) from Sweet Apple Acres in Spring vs. Fall for the last 10 years.
      - Comparing the "Wonderbolts Tryouts" performance scores for Spitfire and Skyflare (twins).
      
## Independent vs. Dependent Data      

- Are the following dependent or independent?
    1. Rainbow Dash times two separate groups, Pegasi trainees and Unicorn cadets, on the same 200-meter aerial course. 
    2. Twilight Sparkle measures her own spell‐casting accuracy before and after attending Princess Celestia's advanced magic workshop.  
    3. Applejack records bushel counts from Sweet Apple Acres in spring this year and compares them to bushel counts from Sugarcube's orchard over the same period.  
    4. The Cutie Mark Crusaders each take a friendship-lesson quiz, and their scores are compared to a completely different group of ponies at the School of Friendship.  
    5. Fluttershy records the heart rates of the same group of critters before and after she plays soothing music for them.  


## Confidence Intervals: Two Independent Means

$(1-\alpha)100\%$ confidence interval for $\mu_1-\mu_2$:

$$
(\bar{x}_1 - \bar{x}_2) \pm t_{\alpha/2} \sqrt{\frac{s_1^2 }{n_1} + \frac{s_2^2}{n_2}}
$$

- where 
    - $(\bar{x}_1-\bar{x}_2)$ is the point estimate for $\mu_1-\mu_2$
        - $\bar{x}_i$ is the sample mean for group $i$
    - $t_{\alpha/2}$ has $\min(n_1-1, n_2-1)$ degrees of freedom
    - $s_i^2$ is the sample variance for group $i$
    - $n_i$ is the sample size of group $i$

## Confidence Intervals: Two Independent Means (R)

- We will use the `independent_mean_CI` function from `library(ssstats)` to find the confidence interval.

- Generic syntax:

```{r}
#| eval: false
dataset_name %>% independent_mean_CI(grouping = grouping_variable,
                                     continuous = continuous_variable, 
                                     confidence = confidence_level)
```

## Confidence Intervals: Two Independent Means

- The Pegasus trainers insist that a healthy Pony munches through 25 apples per day to stay strong and energetic. Looking for differences between those that are above and below target wing-flap rates, a researcher visits the apple stands at Sweet Apple Acres and records the exact number of apples each of the Pegasi in training eats in a typical day. 

- Use the wing-flap data to estimate the difference in apple consumption (*apples*) betwen those that are above or below the target rate (*target*). Estimate using a 95% confidence interval.

- How should we change the following code?

```{r}
#| eval: false
dataset_name %>% independent_mean_CI(grouping = grouping_variable,
                                     continuous = continuous_variable, 
                                     confidence = confidence_level)
```

## Confidence Intervals: Two Independent Means

- The Pegasus trainers insist that a healthy Pony munches through 25 apples per day to stay strong and energetic. Looking for differences between those that are above and below target wing-flap rates, a researcher visits the apple stands at Sweet Apple Acres and records the exact number of apples each of the Pegasi in training eats in a typical day. 

- Use the wing-flap data to estimate the difference in apple consumption (*apples*) betwen those that are above or below the target rate (*target*). Estimate using a 95% confidence interval.

- Our updated code should look like:

```{r}
#| eval: false
wing_flap %>% independent_mean_CI(grouping = target,
                                  continuous = apples, 
                                  confidence = 0.95)
```

## Confidence Intervals: Two Independent Means

- Running the code,

```{r}
wing_flap %>% independent_mean_CI(grouping = target,
                                  continuous = apples, 
                                  confidence = 0.95)
```

- Thus, the 95% CI for $\mu_{\text{above}} - \mu_{\text{below}}$ is (8.13, 11.98).
    - The pegasi above the target wing-flap rate eat, on average, somewhere between 8 and 12 more apples than those below the target wing-flap rate.
    
## Hypothesis Testing: Two Independent Means

- **Hypotheses: Two Tailed**
    - $H_0: \ \mu_1-\mu_2=\mu_0$
    - $H_1: \ \mu_1-\mu_2 \ne \mu_0$
    
- **Hypotheses: Left Tailed**    
    - $H_0: \ \mu_1-\mu_2 \ge \mu_0$
    - $H_1: \ \mu_1-\mu_2 < \mu_0$
    
- **Hypotheses: Right Tailed**    
    - $H_0: \ \mu_1-\mu_2 \le \mu_0$
    - $H_1: \ \mu_1-\mu_2 > \mu_0$  

## Hypothesis Testing: Two Independent Means

**Test Statistic:**

$$
t_0 = \frac{(\bar{x}_1-\bar{x}_2)-\mu_0}{{\sqrt{\frac{s_1^2}{n_1} + \frac{s^2_2}{n_2}}}}
$$

- where
    - $\bar{x}_i$ is the mean for group $i$
    - $\mu_0$ is the hypothesized difference
    - $s_i^2$ is the sample variance for group $i$
    - $n_i$ is the sample size of group $i$    
    - $\text{df} = \text{min}(n_1-1, n_2-1)$
    
## Hypothesis Testing: Two Independent Means

***p*-Value:**

- **$p$-value: Two Tailed**

$$p = 2\times P\left[t_{\text{df}} \ge |t_0|\right]$$

- **$p$-value: Left Tailed**    

$$p = P\left[t_{\text{df}} \le t_0\right]$$
    
- **$p$-value: Right Tailed**    

$$p = P\left[t_{\text{df}} \ge t_0\right]$$

## Hypothesis Testing: Two Independent Means (R)

- We will use the `independent_mean_HT` function from `library(ssstats)` to perform the necessary calculations for the hypothesis test.

- Generic syntax:

```{r}
#| eval: false
dataset_name %>% independent_mean_HT(grouping = grouping_variable,
                                     continuous = continuous_variable, 
                                     mu = hypothesized_value, 
                                     alternative = "alternative_direction", 
                                     alpha = specified_alpha)
```

- For the entered variable (*continuous*), we will see:
    - Hypotheses (based on *hypothesized_value* and *alternative*)
    - Test statistic and p-value
    - Conclusion
    
- **Note!** When looking at the grouping variable, R will subtract in alphabetic/numeric order. 

## Hypothesis Testing: Two Independent Means

- Perform the appropriate hypothesis test to determine if the above target pegasi are eating 5 or more apples than the below target pegasi. Test at the $\alpha=0.05$ level.

- What is the direction of the test? How do you know? <br><br><br><br>

- What is the hypothesized value? How do you know? <br><br><br><br>

- What are the corresponding hypotheses?

## Hypothesis Testing: Two Independent Means

- Perform the appropriate hypothesis test to determine if the above target pegasi are eating 5 or more apples than the below target pegasi. Test at the $\alpha=0.05$ level.

- How should we change the following code?

```{r}
#| eval: false
dataset_name %>% independent_mean_HT(grouping = grouping_variable,
                                     continuous = continuous_variable, 
                                     mu = hypothesized_value, 
                                     alternative = "alternative_direction", 
                                     alpha = specified_alpha)
```

## Hypothesis Testing: Two Independent Means

- Perform the appropriate hypothesis test to determine if the above target pegasi are eating 5 or more apples than the below target pegasi. Test at the $\alpha=0.05$ level.

- Our updated code should look like:

```{r}
#| eval: false
wing_flap %>% independent_mean_HT(grouping = target,
                                  continuous = apples, 
                                  mu = 5, 
                                  alternative = "greater", 
                                  alpha = 0.05)
```

## Hypothesis Testing: Two Independent Means

- Running the code,

```{r}
wing_flap %>% independent_mean_HT(grouping = target,
                                  continuous = apples, 
                                  mu = 5, 
                                  alternative = "greater", 
                                  alpha = 0.05)
```

## Hypothesis Testing: Two Independent Means

- **Hypotheses:**
    - $H_0: \ \mu_{\text{above}} - \mu_{\text{below}} \le 5$
    - $H_1: \ \mu_{\text{above}} - \mu_{\text{below}} > 5$

- **Test Statistic and *p*-Value**
    - $t_0 = 5.445$, $p < 0.001$
    
- **Rejection Region**
    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$
    
- **Conclusion and interpretation**
    - Reject $H_0$ ($p \text{ vs } \alpha \to p < 0.001 < 0.05$). There is sufficient evidence to suggest that ponies above target, on average, eat 5 more apples than those below target.

## Two Dependent Means: Summary Statistics

-   We are now interested in comparing two **dependent** groups.

-   We assume that the two groups come from the same population and are going to examine the difference,

$$
d = y_{i, 1} - y_{i, 2}
$$

-   After drawing samples, we have the following,
    -   $\bar{d}$ estimates $\mu_d$,
    -   $s^2_d$ estimates $\sigma^2_d$, and
    -   $n$ is the sample size.
    
## Two Dependent Means: Summary Statistics (R)  

- We will use the `dependent_mean_median` function from `library(ssstats)` to find the summary statistics for this data.

- Generic syntax:

```{r}
#| eval: false
dataset_name %>% dependent_mean_median(col1 = first_variable,
                                       col2 = second_variable)
```

- Note that this will compute summary statistics for:
    - $x_d = x_1-x_2$
    - $x_1$
    - $x_2$

## Two Dependent Means: Summary Statistics

- Princess Celestia has invited two groups of flyers to take part in a brand-new "SkyStride" aerial training camp. Before the camp begins, each pony perches on a floating platform while a team of Wonderbolt engineers use magical sensors to record their baseline wing-flap rate (flaps per second) as they hover in place (*pre_training_wfr*).

- Over two weeks, trainees attend identical flight drills: precision loops, cloud-weaving obstacle courses, and high-altitude sprints. At camp's end, each flyer returns to the sensor platforms for post-training measurements (*post_training_wfr*).

- Let's find the summary statistics. How should this code be edited?

```{r}
#| eval: false
dataset_name %>% dependent_mean_median(col1 = first_variable,
                                       col2 = second_variable)
```

## Two Dependent Means: Summary Statistics

- Princess Celestia has invited two groups of flyers to take part in a brand-new "SkyStride" aerial training camp. Before the camp begins, each pony perches on a floating platform while a team of Wonderbolt engineers use magical sensors to record their baseline wing-flap rate (flaps per second) as they hover in place (*pre_training_wfr*).

- Over two weeks, trainees attend identical flight drills: precision loops, cloud-weaving obstacle courses, and high-altitude sprints. At camp's end, each flyer returns to the sensor platforms for post-training measurements (*post_training_wfr*).

- Our code is as follows:

```{r}
#| eval: false
wing_flap %>% dependent_mean_median(col1 = pre_training_wfr, 
                                    col2 = post_training_wfr)
```

## Two Dependent Means: Summary Statistics

- Running the code,

```{r}
wing_flap %>% dependent_mean_median(col1 = pre_training_wfr, 
                                    col2 = post_training_wfr)
```

## Confidence Intervals: Two Dependent Means

$\mathbf{(1-\boldsymbol\alpha)100\%}$ confidence interval for $\mathbf{\boldsymbol\mu_d}$

$$ \bar{d} \pm t_{\alpha/2} \frac{s_d}{\sqrt{n}} $$

- where 
    - $\bar{d} = \text{mean}(x_1-x_2)$ is the point estimate for $\mu_d = \mu_1-\mu_2$
    - $t_{\alpha/2}$ has $n-1$ degrees of freedom
    - $s_d$ is the sample standard deviation of the *difference*
    - $n$ is the number of pairs of observations

## Confidence Intervals: Two Dependent Means (R)

- We will use the `dependent_mean_CI` function from `library(ssstats)` to find the confidence interval.

- Generic syntax:

```{r}
#| eval: false
dataset_name %>% dependent_mean_CI(col1 = first_group,
                                   col2 = second_group,
                                   confidence = confidence_level)
```


## Confidence Intervals: Two Dependent Means

- We now want to find the 99% CI for the average improvement in wing-flap rate.
    - Hint: improvement can be measured with *post* - *pre*.
    - Hint 2: post measurement: *post_training_wfr*, pre measurement: *pre_training_wfr*.

- How should we change the following code?

```{r}
#| eval: false
dataset_name %>% dependent_mean_CI(col1 = first_group,
                                   col2 = second_group,
                                   confidence = confidence_level)
```

## Confidence Intervals: Two Dependent Means

- We now want to find the 99% CI for the average improvement in wing-flap rate.
    - Hint: improvement can be measured with *post* - *pre*.
    - Hint 2: post measurement: *post_training_wfr*, pre measurement: *pre_training_wfr*.

- Our updated code should look like:

```{r}
#| eval: false
wing_flap %>% dependent_mean_CI(col1 = post_training_wfr,
                                col2 = pre_training_wfr,
                                confidence = 0.99)
```

## Confidence Intervals: Two Dependent Means

- Running the code,

```{r}
wing_flap %>% dependent_mean_CI(col1 = post_training_wfr,
                                col2 = pre_training_wfr,
                                confidence = 0.99)
```

- The 99% confidence interval for $\mu_d$ is (-7.16, 5.33).

## Confidence Intervals: Two Dependent Means

- What happens if we flip the order?

```{r}
#| eval: false
wing_flap %>% dependent_mean_CI(col1 = pre_training_wfr,
                                col2 = post_training_wfr,
                                confidence = 0.99)
```

## Confidence Intervals: Two Dependent Means

- What happens if we flip the order?

```{r}
wing_flap %>% dependent_mean_CI(col1 = pre_training_wfr,
                                col2 = post_training_wfr,
                                confidence = 0.99)
```

- The 99% confidence interval for $\mu_d$ is (-5.33, 7.16).

## Confidence Intervals: Two Dependent Means

- When looking at post - pre, the CI was (-7.16, 5.33).

- When looking at pre - post, the CI was (-5.33, 7.16).

- What is the relationship? <br><br><br><br><br><br>

- Why does the order matter?

## Hypothesis Testing: Two Dependent Means

- **Hypotheses: Two Tailed**
    - $H_0: \ \mu_d=\mu_0$
    - $H_1: \ \mu_d \ne \mu_0$
    
- **Hypotheses: Left Tailed**    
    - $H_0: \ \mu_d \ge \mu_0$
    - $H_1: \ \mu_d < \mu_0$
    
- **Hypotheses: Right Tailed**    
    - $H_0: \ \mu_d \le \mu_0$
    - $H_1: \ \mu_d > \mu_0$  
    
- **Note!** $\mu_d = \mu_1 - \mu_2$    


## Hypothesis Testing: Two Dependent Means

**Test statistic:**

$$
t_0 = \frac{\bar{d}-\mu_0}{\frac{s_d}{\sqrt{n}}} \sim t_{\text{df}}
$$

- where
    - $\bar{d} = \text{mean}(x_1-x_2)$ is the point estimate for $\mu_d = \mu_1-\mu_2$
    - $\mu_0$ is the hypothesized difference
    - $s_d$ is the sample standard deviation of the *difference*
    - $n$ is the number of pairs of observations
    - $\text{df} = n-1$

## Hypothesis Testing: Two Dependent Means

***p*-Value:**

- **$p$-value: Two Tailed**

$$p = 2\times P\left[t_{\text{df}} \ge |t_0|\right]$$

- **$p$-value: Left Tailed**    

$$p = P\left[t_{\text{df}} \le t_0\right]$$
    
- **$p$-value: Right Tailed**    

$$p = P\left[t_{\text{df}} \ge t_0\right]$$


## Hypothesis Testing: Two Dependent Means (R)

- We will use the `dependent_mean_HT` function from `library(ssstats)` to perform the necessary calculations for the hypothesis test.

- Generic syntax:

```{r}
#| eval: false
dataset_name %>% dependent_mean_HT(col1 = first_group,
                                   col2 = second_group,
                                   alternative = "alternative_direction",
                                   mu = hypothesized_diff,
                                   alpha = alpha_level)
```

## Hypothesis Testing: Two Dependent Means

- Perform the appropriate hypothesis test to determine if there is a difference in wing-flap rate pre- and post-training. Test at the $\alpha=0.01$ level.

- What is the direction of the test? How do you know? <br><br><br><br>

- What is the hypothesized value? How do you know? <br><br><br><br>

- What are the corresponding hypotheses?

## Hypothesis Testing: Two Dependent Means

- Perform the appropriate hypothesis test to determine if there is a difference in wing-flap rate pre- and post-training. Test at the $\alpha=0.01$ level.

- How should we change the following code?

```{r}
#| eval: false
dataset_name %>% dependent_mean_HT(col1 = first_group,
                                   col2 = second_group,
                                   alternative = "alternative_direction",
                                   mu = hypothesized_diff,
                                   alpha = alpha_level)
```

## Hypothesis Testing: Two Dependent Means

- Perform the appropriate hypothesis test to determine if there is a difference in wing-flap rate pre- and post-training. Test at the $\alpha=0.01$ level.

- Our updated code should look like:

```{r}
#| eval: false
wing_flap %>% dependent_mean_HT(col1 = post_training_wfr,
                                col2 = pre_training_wfr,
                                alpha = 0.01)
```

## Hypothesis Testing: Two Dependent Means

- Running the code,

```{r}
wing_flap %>% dependent_mean_HT(col1 = post_training_wfr,
                                col2 = pre_training_wfr,
                                alpha = 0.01)
```

## Hypothesis Testing: Two Dependent Means

- **Hypotheses:**
    - $H_0: \ \mu_{d} = 0$, where $\mu_d = \mu_{\text{pre}} - \mu_{\text{post}}$
    - $H_1: \ \mu_d \ne 0$

- **Test Statistic and *p*-Value**
    - $t_0 = -0.859$, $p = 0.399$
    
- **Rejection Region**
    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.01$
    
- **Conclusion and interpretation**
    - Fail to reject $H_0$ ($p \text{ vs } \alpha \to p = 0.399 > 0.05$). There is not sufficient evidence to suggest that training changed wing-flap rates.

## Wrap Up        

- Today's lecture:
    - Independent *t*-test
    - Dependent *t*-test
    
- Next week:
    - Monday/Tuesday: lab
        - Quiz: two-sample means
    - Wednesday/Thursday: lecture
        - Assumptions on *t*-tests
        - Alternatives to *t*-tests
