---
title: "Two-Sample Means"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    self-contained: true
    slide-number: true
    center: false
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    code-annotations: hover
    incremental: false
editor: source
---
        
## Introduction: Topics

- We previously have reviewed estimation and basic statistical inference.
    - Point estimation
    - Confidence intervals
    - Hypothesis testing
    
- Today we will focus on the following scenarios today:
    - Two-sample means, independent data ($\mu_1-\mu_2$)
    - Two-sample means, dependent data ($\mu_d$)
    
## Introduction: Data

- The Wonderbolts have opened a new Reserve Camp, where 30 cadets spend two weeks training for a shot at the main squad.

- On arrival, each cadet is assigned to one of two training squadrons:
    - Thunder Squadron
    - Lightning Squadron

- The two squadrons train separately and compete against each other in a series of trials.

    - Independent data

- Every cadet is also measured once at the start of camp and once again at the end, after two weeks of identical drills.

    - Dependent data


## Introduction: Data

- Eyeballing the dataset,

```{r}
#| echo: false
library(ssstats)
library(tidyverse)

wonderbolt_camp <- read_csv("https://raw.githubusercontent.com/samanthaseals/STA4173Fa26/refs/heads/main/lectures/data/wonderbolt_camp.csv")
head(wonderbolt_camp)
```


    
## Independent Data

::: {.notebook-entry}
::: {.notebook-tab}
independent data
:::
Data is considered **independent** when observations in one group (or sample) do not influence or relate to observations in another group.
:::

- Examples:
    - Comparing the cruising speeds of a random sample of Pegasi vs. a random sample of Unicorns flying a short course.
    - Measuring friendship lesson quiz scores for a group of Cutie Mark Crusaders vs. a group of Wonderbolts Cadets.
    - Examining the graduation rates between Unicorns and Pegasi.
        
## Dependent Data        

::: {.notebook-entry}
::: {.notebook-tab}
dependent data
:::
Data is considered **dependent** when each observation in the first sample is *paired* with exactly one observation in the second sample.
:::

- Examples:
    - Students' magic‐proficiency scores before and after Princess Celestia's advanced spell workshop.
    - Applejack's apple‐yield (in bushels) from Sweet Apple Acres in Spring vs. Fall for the last 10 years.
    - Comparing the "Wonderbolts Tryouts" performance scores for Spitfire and Skyflare (twins).
      
## Independent vs. Dependent Data      

- Are the following dependent or independent?
    1. Rainbow Dash times two separate groups, Pegasi trainees and Unicorn cadets, on the same 200-meter aerial course. 
    2. Twilight Sparkle measures her own spell‐casting accuracy before and after attending Princess Celestia's advanced magic workshop.  
    3. Applejack records bushel counts from Sweet Apple Acres in spring this year and compares them to bushel counts from Sugarcube's orchard over the same period.  
    4. The Cutie Mark Crusaders each take a friendship-lesson quiz, and their scores are compared to a completely different group of ponies at the School of Friendship.  
    5. Fluttershy records the heart rates of the same group of critters before and after she plays soothing music for them.  


## Confidence Intervals: Two Independent Means

$(1-\alpha)100\%$ confidence interval for $\mu_1-\mu_2$:

$$
(\bar{x}_1 - \bar{x}_2) \pm t_{\alpha/2} \sqrt{\frac{s_1^2 }{n_1} + \frac{s_2^2}{n_2}}
$$

- where 
    - $(\bar{x}_1-\bar{x}_2)$ is the point estimate for $\mu_1-\mu_2$
        - $\bar{x}_i$ is the sample mean for group $i$
    - $t_{\alpha/2}$ has $\min(n_1-1, n_2-1)$ degrees of freedom
    - $s_i^2$ is the sample variance for group $i$
    - $n_i$ is the sample size of group $i$
    
## Hypothesis Testing: Two Independent Means

- **Hypotheses: Two Tailed**
    - $H_0: \ \mu_1-\mu_2=\mu_0$
    - $H_1: \ \mu_1-\mu_2 \ne \mu_0$
    
- **Hypotheses: Left Tailed**    
    - $H_0: \ \mu_1-\mu_2 \ge \mu_0$
    - $H_1: \ \mu_1-\mu_2 < \mu_0$
    
- **Hypotheses: Right Tailed**    
    - $H_0: \ \mu_1-\mu_2 \le \mu_0$
    - $H_1: \ \mu_1-\mu_2 > \mu_0$  

## Hypothesis Testing: Two Independent Means

**Test Statistic:**

$$
t_0 = \frac{(\bar{x}_1-\bar{x}_2)-\mu_0}{{\sqrt{\frac{s_1^2}{n_1} + \frac{s^2_2}{n_2}}}} \sim t_{\text{df}}
$$

- where
    - $\bar{x}_i$ is the mean for group $i$
    - $\mu_0$ is the hypothesized difference
    - $s_i^2$ is the sample variance for group $i$
    - $n_i$ is the sample size of group $i$    
    - $\text{df} = \text{min}(n_1-1, n_2-1)$    

## Two Independent Means (`R`)

- We will use the `independent_means` function from `library(ssstats)` to perform statistical inference.

- Generic syntax:

```{r}
#| eval: false
dataset_name |> independent_means( # <1>
  outcome = outcome_variable, # <2>
  grouping = grouping_variable, # <3>
  alternative = "two" | "less" | "greater", # <4>
  mu = hypothesized_difference, # <5>
  alpha = alpha_level # <6>
  ) 
```
1. Pipes your dataset into `independent_means()`.
2. `outcome`: the continuous variable you want to compare between the two groups.
3. `grouping`: the categorical variable that splits your data into the two groups being compared.
4. `alternative`: the direction of the test; default is `two`.
5. `mu`: the hypothesized difference between the two groups under $H_0$; default is 0.
6. `alpha`: the significance level; default is 0.05.

## Example 1

- Every cadet runs a timed sprint course as part of camp trials.

- The coaches want us to determine if the Thunder Squadron and Lightning Squadron are finishing the sprint course at different times.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, squadron, sprint_time_sec) |> head(n=5)
```



## Example 1

- The coaches want us to determine if the Thunder Squadron and Lightning Squadron are finishing the sprint course at different times.

- What is the direction of the test? How do you know? <br><br><br>

- What is the hypothesized value? How do you know? <br><br><br>

- What are the corresponding hypotheses?


## Example 1

- The coaches want us to determine if the Thunder Squadron and Lightning Squadron are finishing the sprint course at different times.

- We will use the `independent_means()` function,

```{r}
#| eval: false
wonderbolt_camp |> independent_means(outcome = sprint_time_sec,
                                     grouping = squadron, 
                                     alternative = "two",
                                     mu = 0,
                                     alpha = 0.05) 
```

## Example 1

- Running the code,

```{r}
#| echo: false
wonderbolt_camp |> independent_means(outcome = sprint_time_sec,
                                     grouping = squadron, 
                                     alternative = "two",
                                     mu = 0,
                                     alpha = 0.05) 
```

## Example 1

- **Hypotheses:**
    - $H_0: \ \mu_{\text{L}} - \mu_{\text{T}} = 0$
    - $H_1: \ \mu_{\text{L}} - \mu_{\text{T}} \ne 0$

- **Test Statistic and *p*-Value**
    - $t_0 = 2.384$, $p = 0.024$
    
- **Rejection Region**
    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$
    
- **Conclusion and interpretation**
    - Reject $H_0$. There is sufficient evidence to suggest that there is a difference between the Lightning and Thunder squadrons.
    
## Example 1

- We are also given summary information from the output.

- In particular, we saw that

$$
\bar{y}_{\text{L}} = 44.46 \ \ \text{\&} \ \ \bar{y}_{\text{T}} = 41.82
$$

- Then, our point estimate for $\mu_{\text{L}} - \mu_{\text{T}}$ is $2.65$.

- How good is this estimate? Let's examine the 95% CI for $\mu_{\text{L}} - \mu_{\text{T}}$:

$$(0.37, 4.92)$$

## Example 2

- At camp, judges score each cadet's aerial loop maneuver on a 0–100 accuracy scale during trials.

- Based on her gut feeling, the head judge believes that, on average, the Thunder Squadron is performing more accurate loop maneuvers.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, squadron, loop_accuracy_pct) |> head(n=5)
```

## Example 2

- Based on her gut feeling, the head judge believes that, on average, the Thunder Squadron is performing more accurate loop maneuvers.

- What is the direction of the test? How do you know? <br><br><br>

- What is the hypothesized value? How do you know? <br><br><br>

- What are the corresponding hypotheses?

## Example 2

- A quick note about `R` and subtraction!

- Remember that we are estimating $\mu_1 - \mu_2$ -- how does `R` determine $1$ and $2$?

- `R` is going to subtract in <u>alphabetical order</u>. 

    - In our case, with "Thunder" & "Lightning", we are estimating $\mu_{\text{L}} - \mu_{\text{T}}$

- If the Thunder Squadron is more accurate $\to$ higher score $\to$

    - $H_0: \mu_{\text{L}} \ge \mu_{\text{T}}$
    - $H_1: \mu_{\text{L}} < \mu_{\text{T}}$
    
- When specifying the direction in `R`, we will use the `less` option.    

## Example 2

- Based on her gut feeling, the head judge believes that, on average, the Thunder Squadron is performing more accurate loop maneuvers.

- We will use the `independent_means()` function,


```{r}
#| eval: false
wonderbolt_camp |> independent_means(outcome = loop_accuracy_pct,
                                     grouping = squadron, 
                                     alternative = "less",
                                     mu = 0,
                                     alpha = 0.05) 
```

## Example 2

- Running the code,

```{r}
#| echo: false
wonderbolt_camp |> independent_means(outcome = loop_accuracy_pct,
                                     grouping = squadron, 
                                     alternative = "less",
                                     mu = 0,
                                     alpha = 0.05) 
```

## Example 2

- **Hypotheses:**
    - $H_0: \ \mu_{\text{L}} - \mu_{\text{T}} \ge 0$
    - $H_1: \ \mu_{\text{L}} - \mu_{\text{T}} < 0$

- **Test Statistic and *p*-Value**
    - $t_0 = -1.255$, $p = 0.111$
    
- **Rejection Region**
    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$
    
- **Conclusion and interpretation**
    - Fail to reject $H_0$. There is sufficient evidence to suggest that there is a difference between the Lightning and Thunder squadrons.
    
## Example 2

- We are also given summary information from the output.

- In particular, we saw that

$$
\bar{y}_{\text{L}} =  78.41 \ \ \text{\&} \ \ \bar{y}_{\text{T}} =  81.29
$$

- Then, our point estimate for $\mu_{\text{L}} - \mu_{\text{T}}$ is $-2.88$.

- How good is this estimate? Let's examine the 95% CI for $\mu_{\text{L}} - \mu_{\text{T}}$:

$$(-7.61, 1.85)$$


## Confidence Intervals: Two Dependent Means

$\mathbf{(1-\boldsymbol\alpha)100\%}$ confidence interval for $\mathbf{\boldsymbol\mu_d}$

$$ \bar{d} \pm t_{\alpha/2} \frac{s_d}{\sqrt{n}} $$

- where 
    - $\bar{d} = \text{mean}(x_1-x_2)$ is the point estimate for $\mu_d = \mu_1-\mu_2$
    - $t_{\alpha/2}$ has $n-1$ degrees of freedom
    - $s_d$ is the sample standard deviation of the *difference*
    - $n$ is the number of pairs of observations

## Hypothesis Testing: Two Dependent Means

- **Hypotheses: Two Tailed**
    - $H_0: \ \mu_d=\mu_0$
    - $H_1: \ \mu_d \ne \mu_0$
    
- **Hypotheses: Left Tailed**    
    - $H_0: \ \mu_d \ge \mu_0$
    - $H_1: \ \mu_d < \mu_0$
    
- **Hypotheses: Right Tailed**    
    - $H_0: \ \mu_d \le \mu_0$
    - $H_1: \ \mu_d > \mu_0$  
    
- **Note!** $\mu_d = \mu_1 - \mu_2$    


## Hypothesis Testing: Two Dependent Means

**Test statistic:**

$$
t_0 = \frac{\bar{d}-\mu_0}{\frac{s_d}{\sqrt{n}}} \sim t_{\text{df}}
$$

- where
    - $\bar{d} = \text{mean}(x_1-x_2)$ is the point estimate for $\mu_d = \mu_1-\mu_2$
    - $\mu_0$ is the hypothesized difference
    - $s_d$ is the sample standard deviation of the *difference*
    - $n$ is the number of pairs of observations
    - $\text{df} = n-1$
    
## Two Dependent Means (`R`)

- We will use the `dependent_means` function from `library(ssstats)` to perform statistical inference.

- Generic syntax:

```{r}
#| eval: false
dataset_name |> dependent_means( # <1>
  col1 = first_variable, # <2>
  col2 = second_variable, # <3>
  alternative = "two" | "less" | "greater", # <4>
  mu = hypothesized_difference, # <5>
  alpha = alpha_level # <6>
  ) 
```
1. Pipes your dataset into `dependent_means()`.
2. `col1`: the first paired measurement in the difference $d = \text{col1} - \text{col2}$.
3. `col2`: the second paired measurement in the difference $d = \text{col1} - \text{col2}$.
4. `alternative`: the direction of the test; default is `two`.
5. `mu`: the hypothesized difference between the two groups under $H_0$; default is 0.
6. `alpha`: the significance level; default is 0.05.

## Example 3

- Each cadet's aerial accuracy is scored when they arrive at camp, and scored again at the end after two weeks of drills.

- Coaches want to know whether cadets are actually flying more accurately by the end of camp.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, accuracy_score_pre, accuracy_score_post) |> head(n=5)
```

## Example 3

- Coaches want to know whether cadets are actually flying more accurately by the end of camp.

- What is the direction of the test? How do you know? <br><br><br>

- What is the hypothesized value? How do you know? <br><br><br>

- What are the corresponding hypotheses?

## Example 3

- Coaches want to know whether cadets are actually flying more accurately by the end of camp.

- We will use the `dependent_means()` function,


```{r}
#| eval: false
wonderbolt_camp |> dependent_means(col1 = accuracy_score_post,
                                   col2 = accuracy_score_pre,
                                   alternative = "greater", 
                                   mu = 0,
                                   alpha = 0.05) 
```

## Example 3

- Running the code,

```{r}
#| echo: false
wonderbolt_camp |> dependent_means(col1 = accuracy_score_post,
                                   col2 = accuracy_score_pre,
                                   alternative = "greater", 
                                   mu = 0,
                                   alpha = 0.05) 
```

## Example 3

- **Hypotheses:**
    - $H_0: \ \mu_{\text{post}} - \mu_{\text{pre}} \le 0$
    - $H_1: \ \mu_{\text{post}} - \mu_{\text{pre}} > 0$

- **Test Statistic and *p*-Value**
    - $t_0 = 7.143$, $p < 0.001$
    
- **Rejection Region**
    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$
    
- **Conclusion and interpretation**
    - Reject $H_0$. There is sufficient evidence to suggest that there is a difference between the Lightning and Thunder squadrons.
    
## Example 3

- We are also given summary information from the output.

- In particular, we saw that our point estimate for $\mu_{\text{post}} - \mu_{\text{pre}}$ is $6.73$.

- How good is this estimate? Let's examine the 95% CI for $\mu_{\text{post}} - \mu_{\text{pre}}$:

$$(4.80, 8.65)$$


## Example 4

- Each cadet's recovery time after a hard trial is timed at the start of camp and again at the end.

- Coaches want to know whether cadets are recovering faster by the end of camp.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, recovery_time_min_pre, recovery_time_min_post) |> head(n=5)
```

## Example 4

- Coaches want to know whether cadets are recovering faster by the end of camp.

- What is the direction of the test? How do you know? <br><br><br>

- What is the hypothesized value? How do you know? <br><br><br>

- What are the corresponding hypotheses?

## Example 4

- Coaches want to know whether cadets are recovering faster by the end of camp.

- We will use the `dependent_means()` function,

```{r}
#| eval: false
wonderbolt_camp |> dependent_means(col1 = recovery_time_min_post,
                                   col2 = recovery_time_min_pre,
                                   alternative = "less", 
                                   mu = 0,
                                   alpha = 0.05) 
```

## Example 4

- Running the code,

```{r}
#| echo: false
wonderbolt_camp |> dependent_means(col1 = recovery_time_min_post,
                                   col2 = recovery_time_min_pre,
                                   alternative = "less", 
                                   mu = 0,
                                   alpha = 0.05) 
```

## Example 4

- **Hypotheses:**
    - $H_0: \ \mu_{\text{post}} - \mu_{\text{pre}} \ge 0$
    - $H_1: \ \mu_{\text{post}} - \mu_{\text{pre}} < 0$

- **Test Statistic and *p*-Value**
    - $t_0 = -5.254$, $p < 0.001$
    
- **Rejection Region**
    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$
    
- **Conclusion and interpretation**
    - Reject $H_0$. There is sufficient evidence to suggest that cadets are recovering faster at the end of camp.
    
## Example 4

- We are also given summary information from the output.

- In particular, we saw that our point estimate for $\mu_{\text{post}} - \mu_{\text{pre}}$ is $-3.26$.

- How good is this estimate? Let's examine the 95% CI for $\mu_{\text{post}} - \mu_{\text{pre}}$:

$$(-4.53, -1.99)$$

## Let's Practice! 1

- Spectators toss bits to cadets who pull off a standout run during trials.

- Camp organizers want to know whether one squadron is out-earning the other in prize bits.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, squadron, prize_bits_won) |> head(n=5)
```

## Let's Practice! 2

- Spectators tip cadets with bits after their runs. The amount tipped is recorded once early in camp and once again near the end.

- Camp organizers want to know whether cadets are earning more in tips as they improve over camp.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, tip_bits_pre, tip_bits_post) |> head(n=5)
```

## Let's Practice! 3

- A sound crew records the peak crowd noise (in decibels) during each cadet's run.

- Camp organizers are curious whether one squadron draws a louder crowd reaction than the other.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, squadron, cheer_decibels) |> head(n=5)
```

## Let's Practice! 4

- The camp medic tracks how many days each cadet spends recovering from an injury over the two-week camp.

- Coaches want to know whether one squadron's training style is leading to more injury days than the other's.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, squadron, injury_days) |> head(n=5)
```

## Let's Practice! 5

- Each cadet's wing-flap rate while hovering in place is measured on arrival, then measured again after two weeks of drills.

- Coaches want to know whether training actually changes a cadet's wing-flap rate.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, wing_flap_rate_pre, wing_flap_rate_post) |> head(n=5)
```

## Let's Practice! 6

- Each cadet rates how confident they feel about their flying on a self-reported scale, once at the start of camp and once at the end.

- Camp staff want to know whether two weeks of training changes how confident cadets feel.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, confidence_rating_pre, confidence_rating_post) |> head(n=5)
```

## Wrap Up        

- Today's lecture:
    - Independent *t*-test
    - Dependent *t*-test
    
- Wednesday: `R` lab -- bring a laptop!
