---
title: "Two-Sample Medians"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    self-contained: true
    slide-number: true
    center: false
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    code-annotations: hover
    incremental: false
editor: source
---
        
## Introduction: Topics

- We previously have reviewed estimation and basic statistical inference.
    - Point estimation
    - Confidence intervals
    - Hypothesis testing
    
- Today we will focus on the following scenarios today:
    - Two-sample medians, independent data ($M_1-M_2$)
    - Two-sample medians, dependent data ($M_d$)
    
## Introduction: Data

- The Wonderbolts have opened a new Reserve Camp, where 30 cadets spend two weeks training for a shot at the main squad.

- On arrival, each cadet is assigned to one of two training squadrons:
    - Thunder Squadron
    - Lightning Squadron

- The two squadrons train separately and compete against each other in a series of trials.

    - Independent data

- Every cadet is also measured once at the start of camp and once again at the end, after two weeks of identical drills.

    - Dependent data


## Introduction: Data

- Eyeballing the dataset,

```{r}
#| echo: false
library(ssstats)
library(tidyverse)

wonderbolt_camp <- read_csv("https://raw.githubusercontent.com/samanthaseals/STA4173Fa26/refs/heads/main/lectures/data/wonderbolt_camp.csv")
head(wonderbolt_camp)
```

## The Normality Assumption

- All *t*-tests assume approximate normality of the data.

    - In the case of one-sample *t*-tests, the measure of interest must somewhat follow a normal distribution.

    - In the case of two-sample *t*-tests, the measure of interest *in each group* must somewhat follow a normal distribution.
    
- Note that a paired *t*-test is technically a one-sample *t*-test, so we will examine normality of the difference. 

## Normality Assumption: Quantile-Quantile Plots

- There are formal tests for normality ([see article here](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3693611/)), however, we will not use them.\
    
- Instead, we will assess normality using a quantile-quantile (Q-Q) plot.

    - A Q-Q plot helps us visually check if our data follows a specific distribution (here, the normal).

- How do we read Q-Q plots?

    - Each dot represents one observation in our dataset.
    
    - If the data follow a normal distribution, we will observe that the dots fall roughly along a straight diagonal line.
    - We focus on the "middle" of the graph.
    
## Normality Assumption: Quantile-Quantile Plots

```{r}
#| echo: false
#| fig-align: 'center'
library(ssstats)
library(tidyverse)
library(ggpubr)
set.seed(187782)
pegasi_flaps <- tibble(flaps = rnorm(100, mean = 120, sd = 10))
qq <- pegasi_flaps %>% ggplot(aes(sample = flaps)) +
  stat_qq_line(color = "gray", linewidth = 1) +
  stat_qq() +
  labs(x = "Theoretical Quantiles",
       y = "Sample Quantiles",
       title = "Quantile-Quantile Plot") +
  theme_bw()

hist <- pegasi_flaps %>% ggplot(aes(x = flaps)) +
  geom_histogram(fill = "gray", color = "black", bins = 10) +
  labs(title = "Histogram", x = "") + 
  theme_bw()

ggarrange(qq, hist, ncol = 2)
```

## Normality Assumption: Quantile-Quantile Plots

```{r}
#| echo: false
#| fig-align: 'center'
set.seed(881)
pegasi_flaps <- tibble(flaps = rexp(100, rate = 1/50))
qq <- pegasi_flaps %>% ggplot(aes(sample = flaps)) +
  stat_qq_line(color = "gray", linewidth = 1) +
  stat_qq() +
  labs(x = "Theoretical Quantiles",
       y = "Sample Quantiles",
       title = "Quantile-Quantile") +
  theme_bw()

hist <- pegasi_flaps %>% ggplot(aes(x = flaps)) +
  geom_histogram(fill = "gray", color = "black", bins = 10) +
  labs(title = "Histogram", x = "") + 
  theme_bw()

ggarrange(qq, hist, ncol = 2)
```

## Normality Assumption: Quantile-Quantile Plots

```{r}
#| echo: false
#| fig-align: 'center'
set.seed(188277)
pegasi_flaps <- tibble(flaps = rt(100, df = 3) * 10 + 50)
qq <- pegasi_flaps %>% ggplot(aes(sample = flaps)) +
  stat_qq_line(color = "gray", linewidth = 1) +
  stat_qq() +
  labs(x = "Theoretical Quantiles",
       y = "Sample Quantiles",
       title = "Quantile-Quantile") +
  theme_bw()


hist <- pegasi_flaps %>% ggplot(aes(x = flaps)) +
  geom_histogram(fill = "gray", color = "black", bins = 10) +
  labs(title = "Histogram", x = "") + 
  theme_bw()

ggarrange(qq, hist, ncol = 2)
```


## Normality Assumption: Quantile-Quantile Plots

```{r}
#| echo: false
#| fig-align: 'center'
set.seed(658236)
pegasi_flaps <- tibble(flaps = rbeta(100, shape1 = 5, shape2 = 1) * 100)
qq <- pegasi_flaps %>% ggplot(aes(sample = flaps)) +
  stat_qq_line(color = "gray", linewidth = 1) +
  stat_qq() +
  labs(x = "Theoretical Quantiles",
       y = "Sample Quantiles",
       title = "Quantile-Quantile") +
  theme_bw()

hist <- pegasi_flaps %>% ggplot(aes(x = flaps)) +
  geom_histogram(fill = "gray", color = "black", bins = 10) +
  labs(title = "Histogram", x = "") + 
  theme_bw()

ggarrange(qq, hist, ncol = 2)
```

## Normality Assumption (`R`)

- We will check this graphically.

- For independent data,

```{r}
#| eval: false
dataset_name |> independent_qq(outcome = outcome_variable,
                               grouping = grouping_variable)
```

- For dependent data,

```{r}
#| eval: false
dataset_name |> dependent_qq(col1 = first_measurement,
                             col2 = second_measurement)
```

## Example 1

- Every cadet runs a timed sprint course as part of camp trials.

- The coaches want us to determine if the Thunder Squadron and Lightning Squadron are finishing the sprint course at different times.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, squadron, sprint_time_sec) |> head(n=5)
```

## Example 1

- Checking the normality assumption,

```{r}
#| fig-align: 'center'
wonderbolt_camp |> independent_qq(outcome = sprint_time_sec,
                                  grouping = squadron)
```

## Example 2

- At camp, judges score each cadet's aerial loop maneuver on a 0–100 accuracy scale during trials.

- Based on her gut feeling, the head judge believes that, on average, the Thunder Squadron is performing more accurate loop maneuvers.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, squadron, loop_accuracy_pct) |> head(n=5)
```

## Example 2

- Checking the normality assumption,

```{r}
#| fig-align: 'center'
wonderbolt_camp |> independent_qq(outcome = loop_accuracy_pct,
                                  grouping = squadron)
```

## Example 3

- Each cadet's aerial accuracy is scored when they arrive at camp, and scored again at the end after two weeks of drills.

- Coaches want to know whether cadets are actually flying more accurately by the end of camp.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, accuracy_score_pre, accuracy_score_post) |> head(n=5)
```

## Example 3

- Checking the normality assumption,

```{r}
#| fig-align: 'center'
wonderbolt_camp |> dependent_qq(col1 = accuracy_score_pre,
                                col2 = accuracy_score_post)
```

## Example 4

- Each cadet's recovery time after a hard trial is timed at the start of camp and again at the end.

- Coaches want to know whether cadets are recovering faster by the end of camp.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, recovery_time_min_pre, recovery_time_min_post) |> head(n=5)
```

## Example 4

- Checking the normality assumption,

```{r}
#| fig-align: 'center'
wonderbolt_camp |> dependent_qq(col1 = recovery_time_min_pre,
                                col2 = recovery_time_min_post)
```

## Introduction: Two Independent Medians

- The Wilcoxon Rank Sum test is a nonparametric alternative to the two-sample *t*-test.

- Instead of comparing group means, we will now turn to comparing the ranks of the data.

- Let us first consider a simple example, $x: \ 1, 7, 10, 2, 6, 8$

- Our first step is to reorder the data: $x: \ 1, 2, 6, 7, 8, 10$

- Then, we replace with the ranks: $R: \ 1, 2, 3, 4, 5, 6$

## Two Independent Medians: Ranking Data

- What if all data values are not unique? We will assign the <u>average</u> rank for that group.

- For example, $x: 9, 8, 8, 0, 3, 4, 4, 8$

- Let's reorder: $x: 0, 3, 4, 4, 8, 8, 8, 9$

- Rank ignoring ties: $R: 1, 2, 3, 4, 5, 6, 7, 8$

- Now, the final rank: $R: 1, 2, 3.5, 3.5, 6, 6, 6, 8$

## Hypothesis Testing: Two Independent Medians

- **Hypotheses: Two Tailed**
    - $H_0: \ M_1-M_2 = M_0$
    - $H_1: \ M_1-M_2 \ne M_0$
    
- **Hypotheses: Left Tailed**    
    - $H_0: \ M_1-M_2 \ge M_0$
    - $H_1: \ M_1-M_2 < M_0$
    
- **Hypotheses: Right Tailed**    
    - $H_0: \ M_1-M_2 \le M_0$
    - $H_1: \ M_1-M_2 > M_0$  
    
## Hypothesis Testing: Two Independent Medians

$$
T_0 = \sum R_{\text{1}} - \frac{n_1(n_1+1)}{2}
$$

- $T = \sum R_{\text{sample 1}} - \frac{n_1(n_1+1)}{2}$

- where
    - $\sum R_1$ is the sum of the ranks for the first group
    - $n_1$ is the sample size of the first group 
    
- Note that $p =$ (calculated by R :))    

## Two Independent Medians (R)

- We will use the `independent_medians` function from `library(ssstats)` to perform the necessary calculations for the hypothesis test.

- Generic syntax:

```{r}
#| eval: false
dataset_name %>% independent_medians(continuous = continuous_variable,
                                     grouping = grouping_variable,
                                     alternative = "two" | "greater" | "less",
                                     m = hypothesized_difference,
                                     alpha = specified_alpha)
```


## Example 1

- The coaches want us to determine if the Thunder Squadron and Lightning Squadron are finishing the sprint course at different times.

- We will use the `independent_medians()` function,

```{r}
#| eval: false
wonderbolt_camp |> independent_medians(outcome = sprint_time_sec,
                                       grouping = squadron, 
                                       alternative = "two",
                                       m = 0,
                                       alpha = 0.05) 
```

## Example 1

- Running the code,

```{r}
#| echo: false
wonderbolt_camp |> independent_medians(outcome = sprint_time_sec,
                                     grouping = squadron, 
                                     alternative = "two",
                                     m = 0,
                                     alpha = 0.05) 
```

## Example 1

- **Hypotheses:**
    - $H_0: \ M_{\text{L}} - M_{\text{T}} = 0$
    - $H_1: \ M_{\text{L}} - M_{\text{T}} \ne 0$

- **Test Statistic and *p*-Value**
    - $W_0 = 168$, $p = 0.019$
    
- **Rejection Region**
    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$
    
- **Conclusion and interpretation**
    - Reject $H_0$. There is sufficient evidence to suggest that there is a difference between the Lightning and Thunder squadrons.
    
## Introduction: Two Dependent Medians

- The Wilcoxon Signed Rank test is a nonparametric alternative to the dependent *t*-test.

- Instead of examining the mean of the difference, we will now turn to examining the ranks of the differences.

- Before ranking, we will find the difference between the paired observations and eliminate any 0 differences.

    - Note 1: elimniating 0 differences is the big difference between the other tests! 
    - Note 2: because we are eliminating 0 differences, this means that our sample size will update to the number of pairs with a non-0 difference.
    
## Two Dependent Medians: Ranking Data    
    
- When ranking, we the differences are ranked based on the absolute value of the difference.

- Then, ranks can be identified as "positive" or "negative" based on the direction of the difference.

| *X* | *Y* | *D* | \|*D*\| | Rank |
|-----|-----|-----|---------|------|
| 5   | 8   | -3  | 3       | $-$ 1.5 |
| 8   | 5   | 3   | 3       | $+$ 1.5  |
| 4   | 4   | 0   | 0       | ---------   |

- In this (very basic) example, we started with $n=3$, but reduced to $n=2$ due to the 0 difference of the third observation.    

## Hypothesis Testing: Two Dependent Medians

- **Hypotheses: Two Tailed**
    - $H_0: \ M_d=M_0$
    - $H_1: \ M_d \ne M_0$
    
- **Hypotheses: Left Tailed**    
    - $H_0: \ M_d \ge M_0$
    - $H_1: \ M_d < M_0$
    
- **Hypotheses: Right Tailed**    
    - $H_0: \ M_d \le M_0$
    - $H_1: \ M_d > M_0$  
    
- **Note!** $M_d = M_1 - M_2$    

## Hypothesis Testing: Two Dependent Medians

- **Test Statistic**

$$
T_0 =
\begin{cases}
  R_+ = \text{sum of positive ranks} & \text{if left-tailed} \\
  R_- = \text{sum of negative ranks} & \text{if right-tailed} \\
  \min(R_+, R_-) & \text{if two-tailed}
\end{cases}
$$

## Two Dependent Medians (`R`)

- We will use the `dependent_medians` function from `library(ssstats)` to perform statistical inference.

- Generic syntax:

```{r}
#| eval: false
dataset_name |> dependent_medians(col1 = first_variable, # <1>
                                  col2 = second_variable, # <2>
                                  alternative = "two" | "less" | "greater", # <3>
                                  m = hypothesized_difference, # <4>
                                  alpha = alpha_level) # <5>
```
2. `col1`: the first paired measurement in the difference $d = \text{col1} - \text{col2}$.
3. `col2`: the second paired measurement in the difference $d = \text{col1} - \text{col2}$.
4. `alternative`: the direction of the test; default is `two`.
5. `m`: the hypothesized difference between the two groups under $H_0$; default is 0.
6. `alpha`: the significance level; default is 0.05.

## Example 4

- Each cadet's recovery time after a hard trial is timed at the start of camp and again at the end.

- Coaches want to know whether cadets are recovering faster by the end of camp.

- We will use the `dependent_means()` function,

```{r}
#| eval: false
wonderbolt_camp |> dependent_medians(col1 = recovery_time_min_post,
                                     col2 = recovery_time_min_pre,
                                     alternative = "less", 
                                     m = 0,
                                     alpha = 0.05) 
```

## Example 4

- Running the code,

```{r}
#| echo: false
wonderbolt_camp |> dependent_medians(col1 = recovery_time_min_post,
                                   col2 = recovery_time_min_pre,
                                   alternative = "less", 
                                   m = 0,
                                   alpha = 0.05) 
```

## Example 4

- **Hypotheses:**
    - $H_0: \ M_{\text{post}} - M_{\text{pre}} \ge 0$
    - $H_1: \ M_{\text{post}} - M_{\text{pre}} < 0$

- **Test Statistic and *p*-Value**
    - $T_0 = 0$, $p < 0.001$
    
- **Rejection Region**
    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$
    
- **Conclusion and interpretation**
    - Reject $H_0$. There is sufficient evidence to suggest that cadets are recovering faster at the end of camp.
    

## Let's Practice! 1

- Spectators toss bits to cadets who pull off a standout run during trials.

- Camp organizers want to know whether one squadron is out-earning the other in prize bits.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, squadron, prize_bits_won) |> head(n=5)
```

## Let's Practice! 2

- Spectators tip cadets with bits after their runs. The amount tipped is recorded once early in camp and once again near the end.

- Camp organizers want to know whether cadets are earning more in tips as they improve over camp.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, tip_bits_pre, tip_bits_post) |> head(n=5)
```

## Let's Practice! 3

- A sound crew records the peak crowd noise (in decibels) during each cadet's run.

- Camp organizers are curious whether one squadron draws a louder crowd reaction than the other.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, squadron, cheer_decibels) |> head(n=5)
```

## Let's Practice! 4

- The camp medic tracks how many days each cadet spends recovering from an injury over the two-week camp.

- Coaches want to know whether one squadron's training style is leading to more injury days than the other's.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, squadron, injury_days) |> head(n=5)
```

## Let's Practice! 5

- Each cadet's wing-flap rate while hovering in place is measured on arrival, then measured again after two weeks of drills.

- Coaches want to know whether training actually changes a cadet's wing-flap rate.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, wing_flap_rate_pre, wing_flap_rate_post) |> head(n=5)
```

## Let's Practice! 6

- Each cadet rates how confident they feel about their flying on a self-reported scale, once at the start of camp and once at the end.

- Camp staff want to know whether two weeks of training changes how confident cadets feel.

- Looking at the data,

```{r}
#| echo: false
wonderbolt_camp |> select(cadet_name, confidence_rating_pre, confidence_rating_post) |> head(n=5)
```

## Wrap Up        

- Today's lecture:
    - Independent Wilcoxon (rank sum)
    - Dependent Wilcoxon (signed rank)
    
- Monday: `R` lab -- bring a laptop!
