---
title: "Posthoc Testing in One-Way Analyses"
execute:
  echo: true
  warning: false
  message: false
  error: true
format:
  revealjs:
    theme: simple
    css: style_lecture.css
    self-contained: true
    slide-number: true
    center: false
    width: 1600
    height: 900
    df-print: kable
    html-math-method: katex
    code-fold: false
    code-tools: false
    code-annotations: hover
    incremental: false
editor: source
---
  
## Introduction
  
- Previous: one sample, two samples

- Now: more than two groups, continuous outcome

- One-way ANOVA

- Kruskal-Wallis

- posthoc testing

- This lecture: posthoc testing

## `R` Setup 

```{r}
library(tidyverse)
library(ssstats)
apple_harvest <- read_csv("https://raw.githubusercontent.com/samanthaseals/STA4173Fa26/refs/heads/main/lectures/data/apple_harvest.csv")
baltimare_harbor <- read_csv("https://raw.githubusercontent.com/samanthaseals/STA4173Fa26/refs/heads/main/lectures/data/baltimare_harbor.csv")
boutique_gems <- read_csv("https://raw.githubusercontent.com/samanthaseals/STA4173Fa26/refs/heads/main/lectures/data/boutique_gems.csv")
klugetown_bazaar <- read_csv("https://raw.githubusercontent.com/samanthaseals/STA4173Fa26/refs/heads/main/lectures/data/klugetown_bazaar.csv")
ponyville_diner_tips <- read_csv("https://raw.githubusercontent.com/samanthaseals/STA4173Fa26/refs/heads/main/lectures/data/ponyville_diner_tips.csv")
wonderbolts_laptimes <- read_csv("https://raw.githubusercontent.com/samanthaseals/STA4173Fa26/refs/heads/main/lectures/data/wonderbolts_laptimes.csv")
```

## Posthoc Testing

- Recall our hypotheses in one-way ANOVA,

    - $H_0: \mu_1 = \mu_2 = ... = \mu_k$
    - $H_1:$ at least one $\mu_i$ is different

- The $F$ test does not tell us *which* mean is different... only that a difference exists.

- In theory, we could perform repeated $t$ tests to determine pairwise differences.
    
    - Recall that ANOVA is an extension of the $t$ test... or that the $t$ test is a special case of ANOVA.

- However, this will increase the Type I error rate ($\alpha$).

## Posthoc Testing

- Recall that the Type I error rate, $\alpha$, is the probability of *incorrectly* rejecting $H_0$.

    - i.e., we are saying there is a difference between the means when there is actually *not* a difference.

- Suppose we are comparing 5 groups.

    - This is 10 pairwise comparisons!! 
  
        - 1-2, 1-3, 1-4, 1-5, 2-3, 2-4, 2-5, 3-4, 3-5, 4-5

- If we perform repeated $t$ tests under $\alpha=0.05$, we are inflating the Type I error to 0.40! 😵

## Posthoc Testing

- When performing posthoc comparisons, we can choose one of two paths:
  
    - Control the Type I (familywise) error rate.
    
    - Do not control the Type I error rate.

- Note that controlling the Type I error rate is more conservative than when we do not control it.

    - "Conservative" = more difficult to reject.

- Generally, statisticians:
  
    - *do not* control the Type I error rate if examining the results of pilot/preliminary studies that are exploring for general relationships.

    - *do* control the Type I error rate if examining the results of confirmatory studies and are attempting to confirm relationships observed in pilot/preliminary studies.

## Posthoc Testing

- The posthoc tests we will learn:
  
    - Tukey's test (ANOVA)
    
        - Performs all pairwise tests and controls the Type I error rate
        
    - Fisher's least significant difference (ANOVA)

        - Performs all pairwise tests but does not control the Type I error rate
        
    - Dunn's test (Kruskal-Wallis)
    
        - Performs all pairwise tests; we can choose to control Type I error.

- **Caution**: we should *only* perform posthoc tests if we have determined that a general difference exists!
  
    - i.e., we should have rejected with either ANOVA or Kruskal-Wallis

## Posthoc Testing

- We will be performing *pairwise* comparisons.

- For each test performed, we are testing:
  
    - If using an ANOVA-based method:
  
        - $H_0: \ \mu_i = \mu_j$ 
        - $H_1: \ \mu_i \ne \mu_j$
  
    - If using a KW-based method:
  
        - $H_0: \ M_i = M_j$ 
        - $H_1: \ M_i \ne M_j$   
  
## Example 2: Rarity's Boutique

- Rarity flits between racks of fabric: "Darling, I need your professional opinion. Between my Canterlot Couture line, my everyday Ponyville Casual line, and my absolutely *dazzling* Gala Elegance line, I use wildly different amounts of gemstones. Is the difference really as dramatic as I think?"

```{r}
#| echo: false
boutique_gems |> head(n = 3)
```

## Example 2

- Recall, we decided this data could be analyzed with ANOVA -- we saw a difference between the groups ($p<0.001$).

```{r}
boutique_gems |> anovas(outcome = gem_count,
                        grouping = collection)
```



## Posthoc Testing: Generic (R)

- We will use the `posthoc()` function from `library(ssstats)` to perform Tukey's posthoc test (resulting $p$ are adjusted for multiple comparisons).

```{r}
#| eval: false
dataset_name |> posthoc(outcome = continuous_variable, # <1>
                        grouping = grouping_variable, # <2>
                        adjust = TRUE or FALSE, # <3>
                        nonparametric = TRUE or FALSE, # <4>
                        alpha = alpha_level) # <5>
```
1. `outcome`: the same continuous variable used in your ANOVA or Kruskal-Wallis test.
2. `grouping`: the same grouping variable used in your ANOVA or Kruskal-Wallis test (must have at least 2 levels).
3. `adjust`: whether to correct for running multiple comparisons; default is `TRUE`. For ANOVA, `TRUE` runs Tukey's HSD and `FALSE` runs Fisher's LSD. For Kruskal-Wallis, `TRUE` applies a Bonferroni correction to Dunn's test and `FALSE` leaves it unadjusted.
4. `nonparametric`: `FALSE` for ANOVA-based methods (default); `TRUE` for Kruskal-Wallis based methods.
5. `alpha`: the significance level; default is `0.05`.        
    
## Posthoc Testing: Tukey's Test (ANOVA)

- Tukey's test allows us to do *all* pairwise comparisons while controlling $\alpha$.

- **Hypotheses**

    - $H_0: \ \mu_i = \mu_j$ 
    - $H_1: \ \mu_i \ne \mu_j$
    
- **Test Statistic**

$$
Q = \frac{|\bar{y}_i - \bar{y}_j|}{\sqrt{ \frac{\text{MS}_{\text{E}}}{2} \left( \frac{1}{n_i} + \frac{1}{n_j} \right) }}
$$


## Posthoc Testing: Tukey's (R)

- To use `posthoc()` for Tukey's test,

```{r}
#| eval: false
dataset_name |> posthoc(outcome = continuous_variable, # <1>
                        grouping = grouping_variable, # <2>
                        adjust = TRUE, # <3>
                        nonparametric = FALSE, # <4>
                        alpha = alpha_level) # <5>
```
1. `outcome`: the same continuous variable used in your ANOVA or Kruskal-Wallis test.
2. `grouping`: the same grouping variable used in your ANOVA or Kruskal-Wallis test (must have at least 2 levels).
3. `adjust`: whether to correct for running multiple comparisons; default is `TRUE`. For ANOVA, `TRUE` runs Tukey's HSD and `FALSE` runs Fisher's LSD. For Kruskal-Wallis, `TRUE` applies a Bonferroni correction to Dunn's test and `FALSE` leaves it unadjusted.
4. `nonparametric`: `FALSE` for ANOVA-based methods (default); `TRUE` for Kruskal-Wallis based methods.
5. `alpha`: the significance level; default is `0.05`.   

## Example 2

- Examining Tukey's for our example,

```{r}
boutique_gems |> posthoc(outcome = gem_count,
                         grouping = collection,
                         adjust = TRUE,
                         nonparametric = FALSE)
```

## Posthoc Testing: Fisher's Test (ANOVA)

- Fisher's allows us to test *all* pairwise comparisons but *does not* control the $\alpha$.

- **Hypotheses:**
    - $H_0: \ \mu_i = \mu_j$
    - $H_1: \ \mu_i \ne \mu_j$

- **Test Statistic:**

$$
t = \frac{|\bar{y}_i - \bar{y}_j|}{\sqrt{ \text{MS}_{\text{E}} \left( \frac{1}{n_i} + \frac{1}{n_j} \right) }}
$$

## Posthoc Testing: Fisher's (R)

- To use `posthoc()` for Fisher's test,

```{r}
#| eval: false
dataset_name |> posthoc(outcome = continuous_variable, # <1>
                        grouping = grouping_variable, # <2>
                        adjust = FALSE, # <3>
                        nonparametric = FALSE, # <4>
                        alpha = alpha_level) # <5>
```
1. `outcome`: the same continuous variable used in your ANOVA or Kruskal-Wallis test.
2. `grouping`: the same grouping variable used in your ANOVA or Kruskal-Wallis test (must have at least 2 levels).
3. `adjust`: whether to correct for running multiple comparisons; default is `TRUE`. For ANOVA, `TRUE` runs Tukey's HSD and `FALSE` runs Fisher's LSD. For Kruskal-Wallis, `TRUE` applies a Bonferroni correction to Dunn's test and `FALSE` leaves it unadjusted.
4. `nonparametric`: `FALSE` for ANOVA-based methods (default); `TRUE` for Kruskal-Wallis based methods.
5. `alpha`: the significance level; default is `0.05`.   

## Example 2

- Examining Fisher's for our example,

```{r}
boutique_gems |> posthoc(outcome = gem_count,
                         grouping = collection,
                         adjust = FALSE,
                         nonparametric = FALSE)
```

## Example 1: Sweet Apple Acres

- Applejack has come to us: "Between the Red Delicious row, the Honeycrisp row, and that new Zap Apple row Granny Smith insisted on plantin', I reckon one of 'em oughta be doin' better than the others. Can you tell me if they're really any different, or am I just imaginin' things?"

```{r}
#| echo: false
apple_harvest |> head(n = 3)
```

## Example 1

- Recall, we analyzed the data with the Kruskal-Wallis,

```{r}
apple_harvest |> kruskal_wallis(outcome = bushels,
                                grouping = row)
```

## Example 1

- Halt!!!!

- We do not perform posthoc testing when we **do not** detect a difference at the global level!

- $p = 0.051$ $\to$ FTR $\to$ we did *not* detect a difference in the medians

- **We do not perform posthoc testing when ANOVA or Kruskal-Wallis did *not* show an overall difference.**

- ANOVA and Kruskal-Wallis are omnibus, or global, tests.

    - If the omnibus test does not suggest a difference exists, we do not poke around further.

## Example 4: Baltimare Harbor

- The harbormaster spreads a stack of logbooks across the table: "Four dock crews unload every ship that comes through: Anchor, Tidewater, Saltmarsh, and Driftwood. Some of the sailors have been grumbling that one crew is dragging everyone's schedule behind. Before I say anything to the crews, I want to know if unloading time actually differs across them."

```{r}
#| echo: false
baltimare_harbor |> head(n = 3)
```

## Example 4

- Most students chose to analyze with the Kruskal Wallis,

```{r}
baltimare_harbor |> kruskal_wallis(outcome = unload_minutes,
                                   grouping = crew)
```


## Posthoc Testing: Dunn's Test

- We can also perform posthoc testing in the Kruskal-Wallis setting using Dunn's test.

    - Rather than compare pairwise means, it compares pairwise average ranks.

- **Hypotheses:**
    - $H_0: \ M_{i} = M_{j}$
    - $H_1: \ M_{i} \ne M_{j}$

- **Test Statistic:**

$$
z_0 = \frac{|\bar{R}_i - \bar{R}_j|}{\sqrt{ \frac{n(n+1)}{12} \left( \frac{1}{n_i} + \frac{1}{n_j} \right) }}
$$

## Posthoc Testing: Adjusted Dunn's (R)

- To use `posthoc()` for Dunn's test adjusting for the Type I error rate,

```{r}
#| eval: false
dataset_name |> posthoc(outcome = continuous_variable, # <1>
                        grouping = grouping_variable, # <2>
                        adjust = TRUE, # <3>
                        nonparametric = TRUE, # <4>
                        alpha = alpha_level) # <5>
```
1. `outcome`: the same continuous variable used in your ANOVA or Kruskal-Wallis test.
2. `grouping`: the same grouping variable used in your ANOVA or Kruskal-Wallis test (must have at least 2 levels).
3. `adjust`: whether to correct for running multiple comparisons; default is `TRUE`. For ANOVA, `TRUE` runs Tukey's HSD and `FALSE` runs Fisher's LSD. For Kruskal-Wallis, `TRUE` applies a Bonferroni correction to Dunn's test and `FALSE` leaves it unadjusted.
4. `nonparametric`: `FALSE` for ANOVA-based methods (default); `TRUE` for Kruskal-Wallis based methods.
5. `alpha`: the significance level; default is `0.05`.  

## Example 4

- Examining adjusted Dunn's for our example,

```{r}
baltimare_harbor |> posthoc(outcome = unload_minutes,
                            grouping = crew,
                            adjust = TRUE,
                            nonparametric = TRUE)
```

## Posthoc Testing: Unadjusted Dunn's (R)

- To use `posthoc()` for Dunn's test without adjusting for the Type I error rate,

```{r}
#| eval: false
dataset_name |> posthoc(outcome = continuous_variable, # <1>
                        grouping = grouping_variable, # <2>
                        adjust = FALSE, # <3>
                        nonparametric = TRUE, # <4>
                        alpha = alpha_level) # <5>
```
1. `outcome`: the same continuous variable used in your ANOVA or Kruskal-Wallis test.
2. `grouping`: the same grouping variable used in your ANOVA or Kruskal-Wallis test (must have at least 2 levels).
3. `adjust`: whether to correct for running multiple comparisons; default is `TRUE`. For ANOVA, `TRUE` runs Tukey's HSD and `FALSE` runs Fisher's LSD. For Kruskal-Wallis, `TRUE` applies a Bonferroni correction to Dunn's test and `FALSE` leaves it unadjusted.
4. `nonparametric`: `FALSE` for ANOVA-based methods (default); `TRUE` for Kruskal-Wallis based methods.
5. `alpha`: the significance level; default is `0.05`.  

## Example 4

- Examining unadjusted Dunn's for our example,

```{r}
baltimare_harbor |> posthoc(outcome = unload_minutes,
                            grouping = crew,
                            adjust = FALSE,
                            nonparametric = TRUE)
```

## Wrap Up

- This lecture introduced posthoc testing, the follow-up to ANOVA and Kruskal-Wallis.

    - If ANOVA $\to$ Tukey's (adjusted) or Fisher's (unadjusted)
    
    - If Kruskal-Wallis $\to$ adjusted Dunn's or unadjusted Dunn's
    
- We only perform posthoc testing when the omnibus test (ANOVA or Kruskal-Wallis) has determined there is an overall difference in the means or medians.    

- We also discussed the familywise error rate (FWE), $\alpha$.

    - When we *adjust* to preserve the FWE, we maintain $\alpha = 0.05$, regardless of the number of tests performed.
    
    - When we *do not adjust* to preserve the FWE, we allow $\alpha$ to inflate; the inflation depends on number of pairwise comparisons.
  
